On the uniqueness and stability of dictionaries \\ for sparse representation of noisy signals

### 2019-05-09 manuscripts/IEEE/final_submission.tex
minor

The main result of this work is a very general uniqueness theorem for sparse coding (Thm.~\ref{DeterministicUniquenessTheorem}) directly 
implying (Cor.~\ref{DeterministicUniquenessCorollary}) that stable representations of a dataset $Z$ are guaranteed whenever generating dictionaries $\mathbf{A}$ satisfy a spark condition on supports and the original sparse codes $\mathbf{x}_i$ are sufficiently diverse.  Moreover, we provide an explicit, computable $\varepsilon(\delta_1, \delta_2)$ in (\ref{epsdel}) that is essentially linear in desired accuracy $\delta_1, \delta_2$.

@@

%We delay defining the explicit constant $C_1$ until Section \ref{DUT} (\eqref{Cdef1}).
%To be clear, the implication of Thm.~\ref{DeterministicUniquenessTheorem} is that $Y = \{\mathbf{Ax}_1, \ldots, \mathbf{Ax}_N\}$ has a stable $k$-sparse representation in $\mathbb{R}^m$, with \eqref{def1} guaranteed provided $\varepsilon$ in Def.~\ref{maindef} does not exceed: 
In words, Thm.~\ref{DeterministicUniquenessTheorem} says that the smaller the regularity $r$ of original support hypergraph $\mathcal{H}$ or the difference $\overline m - m$ between assumed and actual number of elements in the latent dictionary, the more columns and coefficients of the original dictionary $\mathbf{A}$ and sparse codes $\mathbf{x}_i$ are guaranteed to be contained (up to noise) in the appropriately labelled and scaled recovered dictionary $\mathbf{B}$ and codes $\mathbf{\overline x}_i$, respectively. 
In the important special case when $\overline m = m$, the theorem directly implies (e.g., Fig.~\ref{noisyrecovery}) that  $Y = \{\mathbf{Ax}_1, \ldots, \mathbf{Ax}_N\}$ has a stable $k$-sparse representation in $\mathbb{R}^m$, with inequalities \eqref{def1} guaranteed in Def.~\ref{maindef} for the following worst-case error $\varepsilon$: 
\begin{align}\label{epsdel}
\varepsilon(\delta_1, \delta_2) := \min \left\{ \frac{\delta_1}{ C_1 }, \frac{ \delta_2 L_{2k}(\mathbf{A})}{ 1 + C_1 \left( \delta_2 + \max_{i \in [N]} \|\mathbf{x}_i\|_1  \right) } \right\}.
\end{align}

@@

In fact, uniqueness guarantees with minimal assumptions apply to all areas of data science and engineering that utilize learned sparse structure. For example, several groups have applied compressed sensing to signal processing tasks; for instance, in MRI analysis \cite{lustig2008compressed}, image compression \cite{Duarte08}, and even the design of an ultrafast camera \cite{Gao14}. It is only a matter of time before these systems incorporate dictionary learning to encode and decode signals (e.g., in a device that learns structure from motion \cite{kong2016prior}), just as scientists have used sparse coding to make sense of their data \cite{jung2001imaging, agarwal2014spatially, lee2016sparse, wu2016stability}.

Assurances such as those offered by our theorems certify that different devices and algorithms will learn equivalent representations given enough data from statistically identical systems.\footnote{To contrast with the current hot topic of ``Deep Learning'', there are few such uniqueness guarantees for these models of data; moreover, even small noise can dramatically alter their output \cite{goodfellow2014explaining}.} Indeed, a main reason for the sustained interest in dictionary learning as an unsupervised method for data analysis seems to be the assumed well-posedness of parameter identification in the sparse coding model, confirmation of which forms the core of our theoretical findings.

We close with some remaining open challenges.  First, there is the basic problem of proving optimal lower bounds on sufficient sample size for uniqueness and stability in sparse coding.  Additionally, given the NP-hardness of verifying the spark condition for matrices, is approximating a solution to Prob.~\ref{InverseProblem} or \ref{OptimizationProblem} also computationally hard?  Finally, as $\ell_1$ relaxations for inferring sparse vectors have, in many cases, the same solutions as for $\ell_0$, is there an $\ell_1$ version of Thm.~\ref{SLCopt}? In what special cases is it efficient to check that a solution to the problems is optimal?

@@

\section{EXTRA}
An additional consequence of Thm.~\ref{DeterministicUniquenessTheorem} of potential practical relevance to is the implied lower bound $L_{2k}(\mathbf{BPD}) \geq L_{2k}(\mathbf{A}) - C_1\varepsilon$ in the case where $\mathbf{A}$ satisfies \eqref{SparkCondition}. Since solving Prob.~\ref{SLCopt} directly requires a combinatorial search over sparse supports for the $\mathbf{\overline x}_i$ for each proposed dictionary $\mathbf{B}$, rendering it intractable in general, a common strategy is to replace the $\ell_0$-norm in \eqref{minsum} with the $\ell_1$-norm, thereby transforming the inference of sparse $\mathbf{\overline x}_i$ for a fixed dictionary $\mathbf{B}$ into a convex optimization solvable by a linear program. A major advance in compressive sensing was the discovery that, for fixed $\mathbf{B}$, solutions to \eqref{minsum} using either norm yield the same sparse vectors $\mathbf{\overline x}_i$ (up to noise) provided $L_{2k}(\mathbf{B})$ is large enough \cite{?}. What our results hint at are conditions on the generating dictionary $\mathbf{A}$ and $k$-sparse codes $\mathbf{x}_i$ under which \emph{all} matrices $\mathbf{B}$ (of bounded column-norm) that solve Prob.~\ref{OptimizationProblem} also share this property by virtue of uniqueness and stability; that is, conditions which guarantee that the sparse codes $\mathbf{\overline x}_i$ solving Prob.~\ref{OptimizationProblem} also solve the convexified problem when the corresponding dictionary $\mathbf{B}$ is held fixed. It remains to determine, however, what practical conditions preclude matrices that don't solve Prob.~\ref{OptimizationProblem} from nonetheless solving the convexified problem.

Regardless, an implication of the above observation is that the following algorithm solves Prob.~\ref{OptimizationProblem} without appeal to combinatorial search:
\begin{enumerate}
\item List all matrices $\mathbf{B}$.
\item For each matrix $\mathbf{B}$, solve $\mathbf{\overline x}_i = \arg \min \|\mathbf{v}\|_1$ s.t. $\|\mathbf{Bv} - \mathbf{z}_i\| \leq \eta$ for $i = 1, \ldots, N$.
\item Quantify the proposed solution $\mathbf{B}$ and $\mathbf{\overline x}_1, \ldots, \mathbf{\overline x}_N$ as $\rho(\mathbf{B}, \mathbf{\overline x}_1, \ldots, \mathbf{\overline x}_N) = \sum_{i=1}^{N} \|\mathbf{\overline x}_i\|_0$.
\item Output $\arg \min_{\mathbf{B}, \mathbf{\overline x}_1, \ldots, \mathbf{\overline x}_N} \left( \rho \right)$.
 \end{enumerate}

>>>>

The main result of this work is a very general uniqueness theorem for sparse coding (Thm.~\ref{DeterministicUniquenessTheorem}) directly 
implying (Cor.~\ref{DeterministicUniquenessCorollary}) that stable representations of a dataset $Z$ are guaranteed whenever generating dictionaries $\mathbf{A}$ satisfy a spark condition on supports and the original sparse codes $\mathbf{x}_i$ are sufficiently diverse (e.g., Fig.~\ref{noisyrecovery}).  Moreover, we provide an explicit, computable $\varepsilon(\delta_1, \delta_2)$ in (\ref{epsdel}) that is linear in desired accuracy $\delta_1, \delta_2$.

@@

%We delay defining the explicit constant $C_1$ until Section \ref{DUT} (\eqref{Cdef1}).
%To be clear, the implication of Thm.~\ref{DeterministicUniquenessTheorem} is that $Y = \{\mathbf{Ax}_1, \ldots, \mathbf{Ax}_N\}$ has a stable $k$-sparse representation in $\mathbb{R}^m$, with \eqref{def1} guaranteed provided $\varepsilon$ in Def.~\ref{maindef} does not exceed: 
In words, Thm.~\ref{DeterministicUniquenessTheorem} says that the smaller the regularity $r$ of original support hypergraph $\mathcal{H}$ or the difference $\overline m - m$ between assumed and actual number of elements in the latent dictionary, the more columns and coefficients of the original dictionary $\mathbf{A}$ and sparse codes $\mathbf{x}_i$ are guaranteed to be contained (up to noise) in the appropriately labelled and scaled recovered dictionary $\mathbf{B}$ and codes $\mathbf{\overline x}_i$, respectively. 
In the important special case when $\overline m = m$, the theorem directly implies that  $Y = \{\mathbf{Ax}_1, \ldots, \mathbf{Ax}_N\}$ has a stable $k$-sparse representation in $\mathbb{R}^m$, with inequalities \eqref{def1} guaranteed in Def.~\ref{maindef} for the following worst-case error $\varepsilon$: 
\begin{align}\label{epsdel}
\varepsilon(\delta_1, \delta_2) := \min \left\{ \frac{\delta_1}{ C_1 }, \frac{ \delta_2 L_{2k}(\mathbf{A})}{ 1 + C_1 \left( \delta_2 + \max_{i \in [N]} \|\mathbf{x}_i\|_1  \right) } \right\}.
\end{align}

@@

In fact, uniqueness guarantees with minimal assumptions apply to all areas of data science and engineering that utilize learned sparse structure. For example, several groups have applied compressed sensing to signal processing tasks; for instance, in MRI analysis \cite{lustig2008compressed}, image compression \cite{Duarte08}, and even the design of an ultrafast camera \cite{Gao14}. It is only a matter of time before these systems incorporate dictionary learning to encode and decode signals (e.g., in a device that learns structure from motion \cite{kong2016prior}), just as scientists have used sparse coding to make sense of their data \cite{jung2001imaging, agarwal2014spatially, lee2016sparse, wu2016stability}. 
Assurances offered by our theorems certify that different devices and algorithms will learn equivalent representations given enough data from statistically identical systems.\footnote{To contrast with the current hot topic of ``Deep Learning'', there are few such uniqueness guarantees for these models of data; moreover, even small noise can dramatically alter their output \cite{goodfellow2014explaining}.} Indeed, a main reason for the sustained interest in dictionary learning as an unsupervised method for data analysis seems to be the assumed well-posedness of parameter identification in the model, confirmation of which forms the core of our findings.

We close with some remaining open challenges.  First, there is the basic problem of proving optimal lower bounds on sufficient sample size for uniqueness and stability in sparse coding.  Additionally, given the NP-hardness of verifying the spark condition for matrices, is approximating a solution to Prob.~\ref{InverseProblem} or \ref{OptimizationProblem} also computationally hard?  Finally, as $\ell_1$ relaxations for inferring sparse vectors have, in many cases, the same solutions as for $\ell_0$, is there an $\ell_1$ version of Thm.~\ref{SLCopt}? In what special cases is it efficient to check that a solution to the problems is optimal? [*** Incorporate this open problem with more words ***]

@@

%\section{EXTRA}
%An additional consequence of Thm.~\ref{DeterministicUniquenessTheorem} of potential practical relevance to is the implied lower bound $L_{2k}(\mathbf{BPD}) \geq L_{2k}(\mathbf{A}) - C_1\varepsilon$ in the case where $\mathbf{A}$ satisfies \eqref{SparkCondition}. Since solving Prob.~\ref{SLCopt} directly requires a combinatorial search over sparse supports for the $\mathbf{\overline x}_i$ for each proposed dictionary $\mathbf{B}$, rendering it intractable in general, a common strategy is to replace the $\ell_0$-norm in \eqref{minsum} with the $\ell_1$-norm, thereby transforming the inference of sparse $\mathbf{\overline x}_i$ for a fixed dictionary $\mathbf{B}$ into a convex optimization solvable by a linear program. A major advance in compressive sensing was the discovery that, for fixed $\mathbf{B}$, solutions to \eqref{minsum} using either norm yield the same sparse vectors $\mathbf{\overline x}_i$ (up to noise) provided $L_{2k}(\mathbf{B})$ is large enough \cite{?}. What our results hint at are conditions on the generating dictionary $\mathbf{A}$ and $k$-sparse codes $\mathbf{x}_i$ under which \emph{all} matrices $\mathbf{B}$ (of bounded column-norm) that solve Prob.~\ref{OptimizationProblem} also share this property by virtue of uniqueness and stability; that is, conditions which guarantee that the sparse codes $\mathbf{\overline x}_i$ solving Prob.~\ref{OptimizationProblem} also solve the convexified problem when the corresponding dictionary $\mathbf{B}$ is held fixed. It remains to determine, however, what practical conditions preclude matrices that don't solve Prob.~\ref{OptimizationProblem} from nonetheless solving the convexified problem.
%
%Regardless, an implication of the above observation is that the following algorithm solves Prob.~\ref{OptimizationProblem} without appeal to combinatorial search:
%\begin{enumerate}
%\item List all matrices $\mathbf{B}$.
%\item For each matrix $\mathbf{B}$, solve $\mathbf{\overline x}_i = \arg \min \|\mathbf{v}\|_1$ s.t. $\|\mathbf{Bv} - \mathbf{z}_i\| \leq \eta$ for $i = 1, \ldots, N$.
%\item Quantify the proposed solution $\mathbf{B}$ and $\mathbf{\overline x}_1, \ldots, \mathbf{\overline x}_N$ as $\rho(\mathbf{B}, \mathbf{\overline x}_1, \ldots, \mathbf{\overline x}_N) = \sum_{i=1}^{N} \|\mathbf{\overline x}_i\|_0$.
%\item Output $\arg \min_{\mathbf{B}, \mathbf{\overline x}_1, \ldots, \mathbf{\overline x}_N} \left( \rho \right)$.
% \end{enumerate}
